### 2022-11-23 ic/aistats2021/paper.tex
Working compile

% \documentclass[twoside]{article}
\documentclass[thesis.tex]{subfiles}

@@

% \runningauthor{Liang, Arora, Tehrani, Li, Tingley, Meijer}
% \begin{abstract}
  In order to construct accurate proposers for Metropolis-Hastings Markov
  Chain Monte Carlo, we integrate ideas from probabilistic graphical models
  and neural networks in a framework we call Lightweight
  Inference Compilation (LIC). LIC implements amortized inference within an
  open-universe declarative probabilistic programming language (PPL).
  Graph neural networks are used to parameterize proposal distributions
  as functions of Markov blankets, which during ``compilation'' are optimized
  to approximate single-site Gibbs sampling distributions. Unlike prior work
  in inference compilation (IC), LIC forgoes importance sampling of linear
  execution traces in favor of operating directly on Bayesian networks.
  Through using a declarative PPL, the Markov blankets of nodes (which may be
  non-static) are queried at inference-time to produce proposers. Experimental
  results show LIC can produce proposers which have less parameters, greater
  robustness to nuisance random variables, and improved posterior sampling in
  a Bayesian logistic regression and $n$-schools inference application.
% \end{abstract}

@@

A reference implementation for LIC and code to reproduce our experiments
have been made publically available \citep{lic2020}.

@@

\Cref{fig:nuisance} compares the results between LIC and trace-based IC
\citep{le2017inference} for this nuisance variable model. Both \texttt{pyprob}'s
defaults (1 layer 4 dimension sample embedding, 64 dimension address
embedding, 8 dimension distribution type embedding, 10 component GMM
proposer, 1 layer 512 dimension LSTM) and LIC's defaults (used for all
experiments in this paper, 1 layer 4 dimension node embedding, 3 layer 8
dimension Markov blanket embedding, 1 layer node proposal network) with a 10
component GMM proposer are trained on 10,000 samples and subsequently used to
draw 100 posterior samples. Although model size is not directly comparable due
differences in model architecture, \texttt{pyprob}'s resulting models were over $7\times$
larger than those of LIC. Furthermore, despite requiring more than $10\times$ longer
time to train, the resulting sampler produced by \texttt{pyprob} yields
an effective sample size almost $5\times$ smaller than that produced by LIC.
\begin{figure}
  \centering
  \begin{tabular}{lccc}
    \toprule
           & \# params & compile time & ESS   \\
    \midrule
    LIC    & 3,358     & 44 sec.      & 49.75 \\
    PyProb & 21,952    & 472 sec.     & 10.99 \\
    \bottomrule
  \end{tabular}
  \caption{Number of parameters, compilation time (10,000 samples), and effective sample size (100 samples)
    for inference compilation in LIC (this work)
    versus \citep{pyprob2020}
  }\label{fig:nuisance}
\end{figure}

>>>>

% \documentclass[twoside]{article}
\documentclass[../../thesis.tex]{subfiles}

@@

% \runningauthor{Liang, Arora, Tehrani, Li, Tingley, Meijer}
% \begin{abstract}

\chapter{Accelerating Metropolis-hastings with lightweight inference compilation}
\label{ch:lic}

While the subsequent chapters depart from a previously common theme of DPPs,
they continue our study of statistical applications of randomized methods.
In this chapter,
we are concerned with the problem of sampling intractable posteriors of
Bayesian graphical models.
In order to construct accurate proposers for Metropolis-Hastings Markov
Chain Monte Carlo, we integrate ideas from probabilistic graphical models
and neural networks in a framework we call Lightweight
Inference Compilation (LIC). LIC implements amortized inference within an
open-universe declarative probabilistic programming language (PPL).
Graph neural networks are used to parameterize proposal distributions
as functions of Markov blankets, which during ``compilation'' are optimized
to approximate single-site Gibbs sampling distributions. Unlike prior work
in inference compilation (IC), LIC forgoes importance sampling of linear
execution traces in favor of operating directly on Bayesian networks.
Through using a declarative PPL, the Markov blankets of nodes (which may be
non-static) are queried at inference-time to produce proposers. Experimental
results show LIC can produce proposers which have less parameters, greater
robustness to nuisance random variables, and improved posterior sampling in
a Bayesian logistic regression and $n$-schools inference application.
Parts of this chapter were originally published in
\fullcite{lic}.
% \end{abstract}

@@

A reference implementation for LIC and code to reproduce our experiments
have been made publically available\footnote{\url{https://github.com/facebookresearch/lightweight-inference-compilation}}.

@@

\Cref{fig:nuisance} compares the results between LIC and trace-based IC
\citep{le2017inference} for this nuisance variable model. Both \texttt{pyprob}'s
defaults (1 layer 4 dimension sample embedding, 64 dimension address
embedding, 8 dimension distribution type embedding, 10 component GMM
proposer, 1 layer 512 dimension LSTM) and LIC's defaults (used for all
experiments in this paper, 1 layer 4 dimension node embedding, 3 layer 8
dimension Markov blanket embedding, 1 layer node proposal network) with a 10
component GMM proposer are trained on 10,000 samples and subsequently used to
draw 100 posterior samples. Although model size is not directly comparable due
differences in model architecture, \texttt{pyprob}'s resulting models were over $7\times$
larger than those of LIC. Furthermore, despite requiring more than $10\times$ longer
time to train, the resulting sampler produced by \texttt{pyprob} yields
an effective sample size almost $5\times$ smaller than that produced by LIC.
\begin{table}
  \centering
  \begin{tabular}{lccc}
    \toprule
           & \# params & compile time & ESS   \\
    \midrule
    LIC    & 3,358     & 44 sec.      & 49.75 \\
    PyProb & 21,952    & 472 sec.     & 10.99 \\
    \bottomrule
  \end{tabular}
  \caption{Number of parameters, compilation time (10,000 samples), and effective sample size (100 samples)
    for inference compilation in LIC (this work)
    versus \citep{pyprob2020}
  }\label{fig:nuisance}
\end{table}

### 2022-11-23 intro.tex
Working compile

% !TeX root = thesis.tex
\chapter{Introduction}

In \cref{ch:design}, we consider approximately optimal Bayesian experimental design
using an adaptive row sampling algorithm based on DPPs and show that it provides
good approximations. Through generalizing the proof techniques \todo{Determinant preserving random matrices}
applied here, in \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
closed-form for the over-parameterized $n < d$ regime and predicts a double-descent phenomenon in
linear regression which closely matches empirical experiments.
In \cref{ch:projections} we isolate the part of the proof involving concentration
of bilinear forms of matrix resolvents away from the DPP-based design in order to obtain bounds on
expected projections $X_S^\dag X_S$ when $X_S = S X$ is obtained by sub-Gaussian sketching matrix $S$.

The second three chapters concern the setting of probabilistic programming, a computational tool
for specifying probabilistic models and automating inference through MCMC and VI. In particular, we
study approximating a target density $p(x)$ with approximations $q(x)$ for the purposes of
importance sampling and variational inference.
In \cref{ch:lic}, we consider parameterizing $q(x)$ by graph neural networks which condition each
node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
\cite{le2017inference} and run-times which depend on sparsity in the graphical model rather than
the length of execution traces.
When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
considers the problem of tail anisotropy through both a theoretical and practical perspective.
We establish that prior fat-tailed estimators \citep{jaini2020tails} are tail isotropic, propose
an anisotropic approximation (fat-tailed variational inference, FTVI) where
an anisotropic product base measures is pushed forwards through a bijective neural network,
and confirm that in practice FTVI improves both density estimation as well as variational inference.
However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
probabilistic program's source code. We define generalized Gamma tail asymptotics for a number of
elementary distributions and establish how the tail asymptotics are transformed under algebraic transformations
such as sums and products. This enables a priori computation of tail parameters, which we show
improves the stability and convergence of a number of inference tasks.

>>>>

\documentclass[thesis.tex]{subfiles}

\begin{document}

\chapter{Introduction}
\label{ch:intro}

When faced with difficult problems, randomized approximations are a tool
commonly employed by quantitative scientists which can offer algernative
algorithms and analyses. In this dissertation, we consider a range of problems
at the intersection of randomized methods and statistics.
Throughout our work, randomized methods are a unifying theme used in the first section to
construct tractable approximations and later as tools for automating computational Bayesian statistics.

Initially in the first three chapters, we develop novel analyses to recent problems
in experimental design, double descent, and random projections by using determinantal
point processes (DPP).
In \cref{ch:design}, we consider approximately optimal Bayesian experimental design
using an adaptive row sampling algorithm based on DPPs and show that it provides
good approximations. Through generalizing the previous chapter's proof techniques,
in \cref{ch:double_descent} an extension of the DPP-based design is analyzed in
closed-form for the over-parameterized $n < d$ regime and predicts a double-descent phenomenon in
linear regression which closely matches empirical experiments.
In \cref{ch:projections} we isolate the part of the proof involving concentration
of bilinear forms of matrix resolvents away from the DPP-based design in order to obtain bounds on
expected projections $X_S^\dag X_S$ when $X_S = S X$ is obtained by sub-Gaussian sketching matrix $S$.
These chapters correspond to the following publications:
\begin{itemize}
    \item \fullcite{bayesian-experimental-design}
    \item \fullcite{surrogate-design}
    \item \fullcite{precise-expressions}
\end{itemize}

In the later chapters we focus on probabilistic programming and developing theory and tools
to automate statistical inference using randomized algorithms based on
Monte Carlo Markoc chain (MCMC) and variational inference (VI).
The second three chapters concern the setting of probabilistic programming, a computational tool
for specifying probabilistic models and automating inference through MCMC and VI. In particular, we
study approximating a target density $p(x)$ with approximations $q(x)$ for the purposes of
importance sampling and variational inference.
In \cref{ch:lic}, we consider parameterizing $q(x)$ by graph neural networks which condition each
node on its Markov blanket. This reduces the conditioning sets for a node, resulting in improvements over
\cite{le2017inference} and run-times which depend on sparsity in the graphical model rather than
the length of execution traces.
When the target $p(x)$ is both multivariate and heavy tailed, \cref{ch:ftvi}
considers the problem of tail anisotropy through both a theoretical and practical perspective.
We establish that prior fat-tailed estimators \citep{jaini2020tails} are tail isotropic, propose
an anisotropic approximation (fat-tailed variational inference, FTVI) where
an anisotropic product base measures is pushed forwards through a bijective neural network,
and confirm that in practice FTVI improves both density estimation as well as variational inference.
However, we find in practice FTVI and other tail-adaptive approximations often have trouble optimizing
the tail parameter. In \cref{ch:hta}, we consider addressing this issue during static analysis of a
probabilistic program's source code. We define generalized Gamma tail asymptotics for a number of
elementary distributions and establish how the tail asymptotics are transformed under algebraic transformations
such as sums and products. This enables a priori computation of tail parameters, which we show
improves the stability and convergence of a number of inference tasks.
These chapters correspond to the following publications:
\begin{itemize}
    \item \fullcite{lic}
    \item \fullcite{ftvi}
    \item \fullcite{gga}
\end{itemize}

\end{document}
